### 2022-06-15 article/main.tex
update paper and appendix

% Title
\title{Vigilantism, Institutions, and Honor Culture:\\ Understanding Attitudes toward Lynching in Brazil\footnote{We thank Rosario Aguilar, Louise Araújo, Nicholas Barnes, Hannah Baron, Dara Kay Cohen, Nicholas Cowen, Daniel J. D'Amico, Miriam Golden, Umberto Mignozzetti, Eduardo Moncada, Irfan Nooruddin, Jonathan Obert, Brian J. Phillips, Catarina Roman, Alexandra Scacco, Livia Schubiger, Natán Skigin, Nicholas Rush Smith, and Georg Vanberg for their valuable comments. We received helpful feedback in seminars at EGAP, the Extra-legal Governance Institute at the University of Oxford, the Kroc-Kellogg Peace, Conflict, Crime and Violence Workshop at Notre Dame, and the Severyns Ravenholt Seminar in Comparative Politics at the University of Washington. This study received IRB approval from Brown University in October 2020 (Protocol 2009002803). Replication materials are available at \url{https://github.com/danilofreire/lynching-experiment-brazil}. We thank the Centre for the Study of Governance \& Society at King's College London and the Templeton Foundation for financial support.}}

% Author
\author{Danilo Freire\footnote{School of Social and Political Sciences, University of Lincoln, \href{mailto:danilofreire@gmail.com}{\texttt{danilofreire@gmail.com}}, \url{https://danilofreire.github.io}.} \and David Skarbek\footnote{Department of Political Science and the Political Theory Project, Brown University, \href{mailto:david_skarbek@brown.edu}{\texttt{david\_skarbek@brown.edu}}, \url{http://davidskarbek.com}.}}

% Date
\date{June 16, 2021}

@@

% Abstract
\begin{abstract}
\doublespacing \noindent Why do people support extrajudicial violence? In this paper, we identify an important role for cultural norms in explaining attitudes toward lynching. In a series of survey experiments with respondents in Brazil, we examine how people view different motivations for lynchings, which characteristics of lynching scenarios garner greater support for lynching, and whether providing different types of information about lynching reduces support for it. We find  that people do not believe that criminal justice institutions should be relied on to solve serious crimes and do believe that in such instances, community members have the right to take vengeance. In particular, our analysis finds that people strongly support the use of extrajudicial violence by families of victims against men who sexually assault and murder women and children. 
\vspace{.25cm}

@@

Although lynchings and mob violence more generally have decreased in advanced democracies, it remains widespread in many parts of the world. Based on US State Department reports, \citet[33]{jung2020lynching} find that between 1976 and 2018, lynchings occurred in all regions of the world and in more than a hundred countries. In particular, Latin America has been severely affected by a recent increase in lynchings. Such a surge in citizen-led violence has reportedly been motivated by growing numbers of drug-related crimes and by police ineffectiveness \citep{mallen2014vigilantes}. Brazil provides a telling example. From 2011 to 2015, the country registered about 2,500 lynching episodes. In 2015, 173 people were killed by angry mobs---nearly one execution every two days \citep{barbara2015vigilantes, oliveira2016mob}. According to José de Souza Martins \citeyearpar{martins2015linchamentos}, who has studied lynchings in Brazil for more than thirty years, these figures are not only the highest in the country's history, but among the highest in the world. The people who participate in lynchings are typically  young men, but they also sometimes include teenage women and girls, elderly women, and even members of the local police \citep{moura2017linchamentos}.\footnote{Brazilian police forces in São Paulo kill hundreds of people every year in a way that looks similar to vigilantism \citep{willis2015killing}. Recent work also documents how organized crime groups in Brazil use extralegal violence in the process of governing favelas extralegally \citep{magaloni2020killing}.}

@@

One obvious downside to survey experiments is the difficulty in making inferences about the external validity of findings. \citet[17-18]{bateson2020politics} suggests that the closer that a fictional survey prompt matches the real world, the better the external validity. Accordingly, our first experiment is based on an actual account of a lynching in Jundia\'{i}, Brazil. In the second experiment, we include a wide range of lynching-scenario characteristics that allow respondents to consider detailed, realistic examples. Our third experiment examines whether providing different types of information can affect respondents' perceptions of lynchings. Our strategy here is similar to that of public marketing campaigns that governments often use to change public opinion. In this experiment, we provide information akin to government messaging to the general public. Many theories of vigilantism presume that public beliefs and public opinion establish norms about lynchings. Even if none of our respondents have participated or will participate in lynchings, they are part of the community that creates norms about lynching.

In the first experiment, we attempt to understand how the underlying motivation for a lynching affects support for it. In particular, we test whether police absence, the slowness of the legal process, and the belief that punishment will not be harsh enough affect support. We find no effect for any treatment. Instead, combined with existing data on measures of police ineffectiveness and state capacity, we argue that the evidence is consistent with the claim that respondents simply do not view the police as a genuine source of assistance for serious crimes. Our second experiment is a conjoint experiment designed to see which characteristics of a lynching lead people to believe it is more or less justified. We find evidence that is consistent with aspects of an honor culture  \citep{nisbett2018culture,grosjean2014history}. In such cultures, people believe that the victims of crimes or their family members have the right to seek revenge, especially for heinous or otherwise serious crimes. People often see crimes such as forcible rape and murder as stains on the social standing of the victim and their family, and they believe it is not the proper role of the state to respond. In the final experiment, we provide information to respondents to attempt to reduce support for lynching. We remind respondents that lynching is a crime, that all citizens have human rights, and that lynchings can prompt violent vendettas. We find large negative effects of reminding respondents that lynching is a crime and that lynchings carry the risk of vendettas. Taken together, these findings are consistent with the argument that people are more likely to support vigilantism when the state cannot be relied on for crime control and when certain aspects of an honor culture are present.

In studying lynching, this  paper contributes to several debates. First, we provide causal estimates of why Brazilians justify lynching, how vigilantes' motivation affects the justification, and how to reduce support for lynchings. Second, we provide evidence that is consistent with the common argument that vigilantism is driven by state absence or weakness. However, state absence or weakness alone is not enough to explain support for vigilantism. Like \cite{jung2020lynching}, we argue that people's beliefs and cultural norms are crucial to whether vigilantism is perceived to be justified. It is not simply the fact that the state is unavailable to enforce the law that leads people to support lynchings. Instead, it is the belief that it is not the proper role of the police to do so and that the response should instead be carried out by the family of the victim. Third, the insight we gain into the relationship between formal and informal enforcement helps us understand political development or lack thereof. Even when legal institutions are present and effective, norms  can undermine people's reliance on them. This is a major challenge in establishing the rule of law in developing countries, where residents often prefer informal solutions to crime over state-based solutions \citep[10]{blair2020peacekeeping}. As a result, the prevalence of lynchings undermines political development and state capacity, which can fall into a self-reinforcing cycle \citep{jung2020lynching}. Solving the problem of discretionary, extrajudicial use of violence is crucial for understanding political and economic development more generally.

\section{Theory and Hypotheses}
\label{sec:theory}

@@

One challenge to the hypothesis in the context of the US frontier is that vigilante activity was also widespread during roughly the same time along the East Coast and in the South, where states were far more established than along the Western frontier. In response, one theory argues that vigilantism in these locations instead arose along ``social frontiers'' \citep{obert2018keeping}. These are places where social relationships and social standing were ambiguous among different racial, ethnic, and class-based groups. If the state was not strong or stable enough to eliminate these ambiguities, then vigilantism emerged as a way to create and entrench social and political identities. Black Americans, in particular, were lynched in shocking numbers and at shocking frequency across the South well into the twentieth century \citep{dray2003hands,seguin2019national}. Likewise, based on studies of executions of Muslim men in Bosnia and a lynching in Jim Crow--era Maryland, Lee Ann Fujii \citeyearpar{fujii2017talk} argues that lynchings serve as ``violent displays'' that send signals to observers in the community and entrench social hierarchies and political identities.

While the racial prejudice that drove violence in the United States is clear, vigilantism more generally need not be driven primarily by racial tensions. For example, \citet{martins2015linchamentos} writes that lynchings in Brazil are not mainly due to racial prejudice. Local accounts report no significant social difference between victims and perpetrators; nor are Black Brazilians particularly targeted by lynch mobs \citep{band2015rocinha, pearson2018latam, sinhoretto2009linchamentos}. These events might be driven by racism, but they do not reproduce the stark, racially disparate patterns that existed in the United States historically \citep[23]{martins2015linchamentos}. \citet[8]{jung2020lynching} likewise note that ``\textit{racist} social control'' is a unique property of American vigilantism. Brazil's patterns thus pose a puzzle for existing explanations for mob violence and stand in sharp contrast to the American experience \citep{smaangs2016doing, wood2011lynching}.

Recent research has shown that the relationship between states and vigilantism is more complicated than some early studies assumed. Based on extensive interviews and fieldwork in post-apartheid South Africa, Nicholas Rush \citet{smith2019contradictions} finds strong support for the idea that people participate in vigilantism because the state is seen to be \textit{too} strong. Residents perceive a strengthening of the state, fuller support for human rights, and a stronger rule of law. As a result of these perceived changes, crime victims and other community members feel that the state will take too long to arrest, charge, and convict a perpetrator. If the perpetrator is released before trial, they might harm others or intimidate witnesses. Finally, people in South Africa fear that any punishment eventually received will not be severe enough. Accordingly, many South Africans believe it is right for people to take justice into their own hands. In a tragic twist of fate, achieving legal reform and a  closer adherence to the rule of law actually sparks greater use of extrajudicial violence. This suggests a second hypothesis:

@@

The studies thus far reviewed tend to focus on state activity. Recent work suggests that residents' perceptions of the state are also important. In a study of Haiti, Danielle F. Jung and Dara Kay Cohen \citeyearpar[3]{jung2020lynching} argue that lynchings are correlated not with state presence, but instead with residents' \textit{views} about the state. They find that perceptions of state legitimacy are far more important in explaining vigilantism than whether state actors are merely present. Residents perceive groups (both state and nonstate actors) as more legitimate if they govern and provide social services. When people view the state as illegitimate because it does not govern, they believe it is their right to participate in vigilantism. This suggests that to explain vigilantism, we must understand both state activity and residents' perceptions of community-state relationships.

Studies of vigilantism in Africa likewise find that culture, norms, and religion are powerful forces shaping the emergence and operation of vigilante groups. In Nigeria, vigilantism takes a variety of shapes, including Hisbah (shari'a enforcement groups), the Bakassi Boys, the Egbusu Boys, and the O'odua People's Congress \citep{pratten2008thief,last2008search, fourchard2008new}. An increase in the influence of neoliberalism (alongside other political changes) also drove vigilantism \citep[5]{pratten2008politics}. In Ghana, nearly 20 percent of the lynchings between 1990 and 2000 happened because people believed that someone (usually an older woman) had used ``mystical powers'' to wound someone's genitals or to cause other health problems  \citep[414]{adinkrah2005vigilante}. Given a relatively weak state, beliefs, culture, and norms shape whether and how vigilante activity happens.

In our study, we find that an honor culture shapes Brazilian beliefs about lynchings. In honor cultures, the reputation and social standing of a person or group is paramount \citep{nisbett2018culture,crowder2017new, vandello2008us, Cao2021honor}. Offenses that threaten to reduce one's standing are seen to give the victim the right (and often the obligation) to respond extralegally \citep{brown2016honor}. Honor cultures are often described as embracing ``a self-protection ethic'' and operate as a system of ``self-help justice'' \citep[1287, 1290]{grosjean2014history}. The offenses need not be serious, as ``everyday challenges, large and small, can undermine an individual's status and reputation, thus rendering his or her honor malleable and uncertain'' \citep[522]{crowder2017new}. However, gendered norms play an important role in honor societies, making heinous crimes especially important to respond to. Being the victim of a serious crime, such as forcible rape, is often seen as tarnishing the reputation of the victim and their family. As a result, as with dueling practices, failure to respond to a harm can affect one's social standing \citep{allen2006duel}. Finally, many honor cultures also follow \textit{lex talionis}, the idea that justice demands an eye for an eye. In places where an honor culture puts the onus to avenge wrongs on local community members, people view state enforcement as less desirable and less legitimate than vigilantism. They see honor as not restored if the state intervenes.

While some of the most prominent studies of honor cultures focus on the history of Scotch-Irish people in the American South \citep[for example,][662-668]{fischer1989albion}, honor cultures still exist in many places today. In the United States, honor cultures exist in some American neighborhoods and are nearly universal in men's prisons across the country \citep{anderson2000code,fleisher2009myth}. Likewise, honor cultures have been a crucial part of clan, tribal, and customary societies in many parts of the world throughout history \citep{weiner2013rule}. As in such societies, it appears that honor cultures tend to emerge in places where people cannot rely on strong and effective states.\footnote{There is a large literature in law and economics that studies how reputation mechanisms can provide order without law \citep{bernstein1992opting,ellickson1991order,greif2006institutions, leeson2014anarchy}.} Cultures of violence can have remarkably long-run effects \citep{vandello2008us}. For example, in the American South, the presence of Scotch-Irish populations in the eighteenth century explains nearly 20 percent of the difference between homicide rates in the North and South today \citep[1286]{grosjean2014history}. An honor culture, and its emphasis on retaliatory norms and violence, did not emerge in the North because the region developed effective formal institutions more quickly. Taken together, the lesson from these studies is that when people cannot rely on state-based institutions to regulate violence, elements of an honor culture can emerge and produce self-help norms, which in turn increase support for vigilantism. This suggests a third hypothesis:

\begin{itemize}
  \item \textbf{Hypothesis 3:} Support for vigilantism will increase if people accept the norms associated with an honor culture.  
  \end{itemize}

In our three experiments, we directly and indirectly tested these hypotheses. We argue that respondents do not believe that state-based legal institutions are the primary way to respond to serious crime. We find that people's attitudes toward vigilantism are shaped strongly by the characteristics of the incident that gives rise to it. In particular, people strongly support lynching by the family of a victim of a sexual offense or murder. Finally, although honor cultures can persist for many decades, our third experiment shows that support for lynchings can be reduced by reminding people of the penalties for lynching and of the risk of sparking vendettas.

@@

The first piece of evidence that the experiment provides concerns how baseline levels of support for lynching vary according to individual characteristics. Table \ref{tab:exp01-baselines} shows the impact of gender, race, and ideology on the subjects' tolerance of lynchings. Our dependent variable ranges from 0 to 100. We observed that men are more likely to support lynchings when compared with women. This is in contrast to research in Uganda, Tanzania, and South Africa, where women support mob vigilantism more than men do because they believe that they are less likely to be subject to false accusations \citep{wilke2021gender}.\footnote{On the relationships between gender and violence, see \citet{kadera2018gendered} and related papers.} In line with \citet{martins2015linchamentos}, we found little evidence that support for lynchings is driven primarily by racial prejudice.\footnote{We recognize that race is socially-constructed and that racial categories can be codified in multiple ways. We adopt the racial classification used by the Brazilian Census Bureau (IBGE) as it is widely employed in social research \citep[e.g.][24]{lapop2018brazil} and intuitive to our respondents. Nevertheless, our findings are also consistent with research that uses different, and often more nuanced, conceptions of race \citep{monk2016consequences, schwartzman2020colour}. } Ideology is strongly correlated with support for lynching, with a negative coefficient for citizens who identify as being on the center left or left.\footnote{The results are very similar when we estimated the models using data from our third experiment. Please refer to Section D.2 in the supplementary material for the more information.}

@@

One interpretation of the null findings for the treatment effects is that state presence and activity have no influence on attitudes toward lynching. If that interpretation is correct, we should reject Hypotheses 1 and 2. However, given the extensive literature discussed above and the qualitative evidence collected in the next experiment (discussed in Section \ref{sec:exp01}), we instead argue that the null results reflect the fact that respondents simply have little to no faith in the criminal legal system in general, and addressing the issues implied in the treatment conditions would have no substantial effect on its perceived effectiveness. Three pieces of evidence support this interpretation. First, many crimes in Brazil are never solved. For example, a recent statistic indicates that the police only solve about 10 percent of the homicides in Brazil, suggesting that even for serious crimes, the police provide little help \citep{pearson2018latam}. Second, Brazilian police forces themselves carry out a substantial number of unjustified killings of people from disadvantaged and minority communities, which undermines their legitimacy and the sense that they can be relied on \citep{willis2015killing}.\footnote{Recent work on Mexico shows that implementing rule of law reforms can reduce police violence \citep{magaloni2020institutionalized}.} In 2019, police killed more than 1,800 people in Rio de Janeiro alone. Third, the Brazilian penal code allows an accused person to appeal each decision several times, so it can take several years (and sometimes more than a decade) before a criminal case is closed \citep{sousa2005utilizaccao}. Given this, people do not believe that a perpetrator will be punished in a timely matter even if put on trial. Finally, along with the rise of Jair Bolsonaro---the far-right president who has publicly called for people to take the law into their own hands \citep{brant2021armas, uol2021bolsonaro}---survey evidence suggests that people have grown considerably more punitive \citep{datafolha2018penademorte}. Taken together, attitudes about lynchings do not seem to be driven by beliefs about \textit{how} the state provides justice. Instead, they reflect the fact that respondents do not, in general, believe that it \textit{can} provide it. In addition, in our next experiment, we found evidence that people who support an honor culture do not think that the state \textit{should} provide justice in certain circumstances.

@@

\begin{figure}[ht]
\includegraphics[width=17cm]{conjoint-black.pdf}
\caption{Relative preferences for lynching-episode attributes (marginal means)}
\centering
\label{fig:exp01}
\end{figure}

@@

The residency of the perpetrator is not a significant driver of support for lynching. Subjects were indifferent as to whether the crime perpetrator lives in the area (0.498, SE = 0.004, $p$-value = 0.668) or in another neighborhood (0.502, SE = 0.004, $p$-value = 0.668). In this sense, it is possible that subjects see lynchings not as a means to protect the area in which they live, which we commonly suppose vigilante groups do, but rather as retribution in specific cases.

The next attribute is type of offense. Rape was by far the most significant attribute in our model (0.719, SE = 0.009, $p$-value $<$ 0.001), followed by murder (0.608, SE = 0.006, $p$-value $<$ 0.001) and molestation (0.538, SE = 0.006, $p$-value $<$ 0.001). It is not surprising that murder was an important offense mentioned by respondents, and the fact that rape appeared as subjects' first choice is not entirely unexpected either. Among incarcerated populations, sex offenders are near-universally seen as having the lowest standing in the community and typically are subject to more violence than most other incarcerated people \citep{fleisher2009myth}. These findings support Hypothesis 3, in that honor cultures tend to view heinous and sex crimes as warranting an extrajudicial response. Overall, we found that respondents do not believe that property crimes, such as car theft (0.351, SE = 0.009, $p$-value $<$ 0.001) and pickpocketing (0.314, SE = 0.006, $p$-value $<$ 0.001), provide justification for lynching. When we analyzed the data by groups, we also saw that poorer respondents, whose household income is less than R\$3,000 (US\$650) per month, are more likely to support the lynching of individuals who commit a property crime.

@@

For age, we observed that respondents were particularly concerned about children as crime victims (0.571, SE = 0.007, $p$-value $<$ 0.001), as this attribute is strongly correlated with support for lynchings. Respondents also thought lynchings are more justified when the victim is a teenager (0.505, SE = 0.007, $p$-value = 0.474), an elderly person (0.479, SE = 0.006, $p$-value $<$ 0.001), or an adult (0.464, SE = 0.006, $p$-value $<$ 0.001). Together, these facts lend credence to the idea that citizens see lynchings as a valid punishment when the victims have little chance of defending themselves.

Lastly, we examined who respondents believed are more legitimate perpetrators of lynchings. Consistent with a key aspect of an honor culture, respondents considered the family to be the most justified in carrying out a lynching (0.534, SE = 0.007, $p$-value $<$ 0.001). Again, the intervals here did not overlap with those of any other category. Bystanders (0.505, SE = 0.007, $p$-value = 0.450), gangs (0.502, SE = 0.007, $p$-value = 0.815), and neighbors (0.492, SE = 0.007, $p$-value = 0.262) do not increase or decrease respondents attitudes about lynching. If the police carry out a lynching, that reduced the perceived justification (0.467, SE = 0.007, $p$-value $<$ 0.001). In sum, respondents did not believe that lynchings should be carried out by the state but did believe they should be used as a tool for individual or family retribution.

>>>>

% Title
\title{Vigilantism and Institutions:\\ Understanding Attitudes toward Lynching in Brazil\footnote{We thank Rosario Aguilar, Louise Araújo, Nicholas Barnes, Hannah Baron, Dara Kay Cohen, Nicholas Cowen, Daniel J. D'Amico, Miriam Golden, Umberto Mignozzetti, Eduardo Moncada, Irfan Nooruddin, Jonathan Obert, Brian J. Phillips, Catarina Roman, Alexandra Scacco, Livia Schubiger, Natán Skigin, Nicholas Rush Smith, and Georg Vanberg for their valuable comments. We received helpful feedback in seminars at EGAP, the Extra-legal Governance Institute at the University of Oxford, the Comparative Workshop at Brown University, the Kroc-Kellogg Peace, Conflict, Crime and Violence Workshop at Notre Dame, and the Severyns Ravenholt Seminar in Comparative Politics at the University of Washington. This study received IRB approval from Brown University in October 2020 (Protocol 2009002803). Replication materials are available at \url{https://github.com/danilofreire/lynching-experiment-brazil}. We thank the Centre for the Study of Governance \& Society at King's College London and the Templeton Foundation for financial support.}}

% Author
\author{Danilo Freire\footnote{School of Social and Political Sciences, University of Lincoln, \href{mailto:danilofreire@gmail.com}{\texttt{danilofreire@gmail.com}}, \url{https://danilofreire.github.io}.} \and David Skarbek\footnote{Department of Political Science, Brown University, \href{mailto:david_skarbek@brown.edu}{\texttt{david\_skarbek@brown.edu}}, \url{http://davidskarbek.com}.}}

% Date
\date{\today}

@@

% Abstract
\begin{abstract}
\doublespacing \noindent Why do people support extrajudicial violence? In a series of survey experiments with respondents in Brazil, we examine how people view different motivations for lynchings, which characteristics of lynching scenarios garner greater support for lynching, and whether providing different types of information about lynching reduces support for it. We find that people do not believe that criminal justice institutions should be relied on to solve serious crimes and do believe that in such instances, community members have the right to take vengeance. In particular, our analysis finds that people strongly support the use of extrajudicial violence by families of victims against men who sexually assault and murder women and children. 
\vspace{.25cm}

@@

Although lynchings and mob violence more generally have decreased in advanced democracies, it remains widespread in many parts of the world. Based on US State Department reports, \citet[33]{jung2020lynching} find that between 1976 and 2018, lynchings occurred in all regions of the world and in more than a hundred countries. In particular, Latin America has been severely affected by a recent increase in lynchings. Such a surge in citizen-led violence has reportedly been motivated by growing numbers of drug-related crimes and by police ineffectiveness \citep{mallen2014vigilantes}. Brazil provides a telling example. From 2011 to 2015, the country registered about 2,500 lynching episodes. In 2015, 173 people were killed by angry mobs---nearly one execution every two days \citep{barbara2015vigilantes, oliveira2016mob}. According to José de Souza Martins \citeyearpar{martins2015linchamentos}, who has studied lynchings in Brazil for more than thirty years, these figures are not only the highest in the country's history, but among the highest in the world. The people who participate in lynchings are typically  young men, but they also sometimes include teenage women and girls, elderly women, and even members of the local police \citep{moura2017linchamentos}.\footnote{Brazilian police forces in São Paulo kill hundreds of people every year in a way that looks similar to vigilantism \citep{willis2015killing}. Recent work also documents how organized crime groups in Brazil use extralegal violence in the process of governing favelas  \citep{magaloni2020killing}.}

@@

One obvious downside to survey experiments is the difficulty in making inferences about the external validity of findings. \citet[17-18]{bateson2020politics} suggests that the closer that a fictional survey prompt matches the real world, the better the external validity. Accordingly, our first experiment is based on an actual account of a lynching in Jundia\'{i}, Brazil. In the second experiment, we include a wide range of lynching-scenario characteristics that allow respondents to consider detailed, realistic examples. Our third experiment examines whether providing different types of information can affect respondents' perceptions of lynchings. Our strategy here is similar to that of public marketing campaigns that governments often use to change public opinion. In this experiment, we provide information akin to government messaging to the general public.

In the first experiment, we attempt to understand how the underlying motivation for a lynching affects support for it. In particular, we test whether police absence, the slowness of the legal process, and the belief that punishment will not be harsh enough affect support. We find no effect for any treatment. Instead, combined with existing data on measures of police ineffectiveness and state capacity, we argue that the evidence is consistent with the claim that respondents simply do not view the police as a genuine source of assistance for serious crimes. Our second experiment is a conjoint experiment designed to see which characteristics of a lynching lead people to believe it is more or less justified. People believe that the victims of crimes or their family members have the right to seek revenge, especially for heinous or otherwise serious crimes. People see crimes such as forcible rape and murder as especially suitable for extrajudicial violence. In the final experiment, we provide information to respondents to attempt to reduce support for lynching. We remind respondents that lynching is a crime, that all citizens have human rights, and that lynchings can prompt violent vendettas. We find large negative effects of reminding respondents that lynching is a crime and that lynchings carry the risk of vendettas.

In studying lynching, this  paper contributes to several debates. First, we provide causal estimates of why Brazilians justify lynching, how vigilantes' motivation affects the justification, and how to reduce support for lynchings. Second, we provide evidence that is consistent with the common argument that vigilantism is driven by state absence or weakness. However, state absence or weakness alone is not enough to explain support for vigilantism. Like \cite{jung2020lynching}, we argue that people's beliefs and cultural norms are crucial to whether vigilantism is perceived to be justified. It is not simply the fact that the state is unavailable to enforce the law that leads people to support lynchings. Instead, it is the belief that it is not the proper role of the police to do so and that the response should instead be carried out by the family of the victim. Third, our findings inform the relationship between political development and informal norms. Even when legal institutions are present, norms  can undermine people's reliance on them. This is a major challenge in establishing the rule of law in developing countries, where residents often prefer informal solutions to crime over state-based solutions \citep[10]{blair2020peacekeeping}. As a result, the prevalence of lynchings undermines political development and state capacity, which can fall into a self-reinforcing cycle \citep{jung2020lynching}. As such, solving the problem of discretionary, extrajudicial use of violence is crucial for understanding political and economic development more generally.

\section{Literature Review and Hypotheses}
\label{sec:theory}

@@

One challenge to the hypothesis in the context of the US frontier is that vigilante activity was also widespread during roughly the same time along the East Coast and in the South, where states were far more established than along the Western frontier. In response, one theory argues that vigilantism in these locations instead arose along ``social frontiers'' \citep{obert2018keeping}. These are places where social relationships and social standing were ambiguous among different racial, ethnic, and class-based groups. If the state was not strong or stable enough to eliminate these ambiguities, then vigilantism emerged as a way to create and entrench social and political identities. Black Americans, in particular, were lynched in shocking numbers and frequency across the South well into the twentieth century \citep{dray2003hands,seguin2019national}. Likewise, based on studies of executions of Muslim men in Bosnia and a lynching in Jim Crow--era Maryland, Lee Ann Fujii \citeyearpar{fujii2017talk} argues that lynchings serve as ``violent displays'' that send signals to observers in the community and entrench social hierarchies and political identities.

While the racial prejudice that drove violence in the United States is clear, vigilantism more generally need not be driven primarily by racial tensions. For example, \citet{martins2015linchamentos} writes that lynchings in Brazil are not mainly due to racial prejudice. Local accounts report no significant social difference between victims and perpetrators; nor are Black Brazilians particularly targeted by lynch mobs \citep{band2015rocinha, pearson2018latam, sinhoretto2009linchamentos}. These events might be driven by racism, but they do not reproduce the stark, racially disparate patterns that existed in the United States historically \citep[23]{martins2015linchamentos}. \citet[8]{jung2020lynching} likewise note that ``\textit{racist} social control'' is a unique property of American vigilantism. Brazil's patterns thus pose a puzzle for existing explanations for mob violence and stand in contrast to the American experience \citep{smaangs2016doing, wood2011lynching}.

Recent research has shown that the relationship between states and vigilantism is more complicated than some early studies assumed. Based on extensive interviews and fieldwork in post-apartheid South Africa, Nicholas Rush \citet{smith2019contradictions} finds that people participate in vigilantism because the state is seen to be \textit{too} strong. Residents perceive a strengthening of the state, fuller support for human rights, and a stronger rule of law. As a result of these perceived changes, crime victims and other community members feel that the state will take too long to arrest, charge, and convict a perpetrator. If the perpetrator is released before trial, they might harm others or intimidate witnesses. Finally, people in South Africa fear that any punishment eventually received will not be severe enough. Accordingly, many South Africans believe it is right for people to take justice into their own hands. As a result, achieving legal reform and a  closer adherence to the rule of law can actually sparks greater use of extrajudicial violence. This suggests a second hypothesis:

@@

Studies of vigilantism in Africa likewise find that religion, norms, and gender are powerful forces shaping the emergence and operation of vigilante groups. For example, in Nigeria, an increase in the influence of neoliberalism (alongside other political changes) drove vigilantism \citep[5]{pratten2008politics}. In Ghana, nearly 20 percent of the lynchings between 1990 and 2000 happened because people believed that someone (usually an older woman) had used ``mystical powers'' to wound someone's genitals or to cause other health problems  \citep[414]{adinkrah2005vigilante}. More generally, gender norms play an important role in the use of extrajudicial violence. The family members of people who are victims of a serious crime, such as forcible rape, often believe it tarnishes the reputation of the victim and their family. As a result, they believe they have the right to punish the offender. This suggests a third hypothesis:

\begin{itemize}
  \item \textbf{Hypothesis 3:} Support for vigilantism will increase when victims are female and their family carries out retribution.  
  \end{itemize}

In our three experiments, we argue that respondents do not believe that state-based legal institutions are the primary way to respond to serious crime. We find that people's attitudes toward vigilantism are shaped strongly by the characteristics of the incident that gives rise to it. In particular, people strongly support lynching by the family of a victim of a sexual offense or murder. Finally, our third experiment shows that support for lynchings can be reduced by reminding people of the penalties for lynching and of the risk of sparking vendettas.

@@

The first piece of evidence that the experiment provides concerns how baseline levels of support for lynching vary according to individual characteristics. Table \ref{tab:exp01-baselines} shows the impact of gender, race, and ideology on the subjects' tolerance of lynchings. Our dependent variable ranges from 0 to 100. We observed that men are more likely to support lynchings when compared with women. This is in contrast to research in Uganda, Tanzania, and South Africa, where women support mob vigilantism more than men do because they believe that they are less likely to be subject to false accusations \citep{wilke2021gender}.\footnote{On the relationships between gender and violence, see \citet{kadera2018gendered} and related papers.} In line with \citet{martins2015linchamentos}, we found little evidence that support for lynchings is driven primarily by outright racial prejudice.\footnote{We recognize that race is socially-constructed and that racial categories can be codified in multiple ways. We adopt the racial classification used by the Brazilian Census Bureau (IBGE) as it is widely employed in social research \citep[e.g.][24]{lapop2018brazil} and intuitive to our respondents. Nevertheless, our findings are also consistent with research that uses different, and often more nuanced, conceptions of race \citep{monk2016consequences, schwartzman2020colour}. } Ideology is strongly correlated with support for lynching, with a negative coefficient for citizens who identify as being on the center left or left.\footnote{The results are very similar when we estimated the models using data from our third experiment. Please refer to Section D.2 in the supplementary material for the more information.}

@@

There were no significant effects on any of the treatments. One possible interpretation of the null findings is that state presence and activity have no influence on attitudes toward lynching. If that interpretation is correct, we should reject Hypotheses 1 and 2.

Given the extensive literature discussed above and the qualitative evidence collected in the next experiment (discussed in Section \ref{sec:exp01}), an alternative explanation is that the null results reflect the fact that respondents simply have little to no faith in the criminal legal system in general. Several pieces of evidence support this interpretation. First, many crimes in Brazil are never solved. For example, a recent statistic indicates that the police only solve about 10 percent of the homicides in Brazil, suggesting that even for serious crimes, the police provide little help \citep{pearson2018latam}. Second, Brazilian police forces themselves carry out a substantial number of unjustified killings of people from disadvantaged and minority communities, which undermines their legitimacy and the sense that they can be relied on \citep{willis2015killing}.\footnote{Recent work on Mexico shows that implementing rule of law reforms can reduce police violence \citep{magaloni2020institutionalized}.} In 2019, police killed more than 1,800 people in Rio de Janeiro alone. Third, the Brazilian penal code allows an accused person to appeal each decision several times, so it can take several years (and sometimes more than a decade) before a criminal case is closed \citep{sousa2005utilizaccao}. Given this, people do not believe that a perpetrator will be punished in a timely matter even if put on trial. Finally, along with the rise of Jair Bolsonaro---the far-right president who has publicly called for people to take the law into their own hands \citep{brant2021armas, uol2021bolsonaro}---survey evidence suggests that people have grown considerably more punitive \citep{datafolha2018penademorte}.

@@

\begin{figure}[ht]
\includegraphics[width=17cm]{conjoint-color.pdf}
\caption{Relative preferences for lynching-episode attributes (marginal means)}
\centering
\label{fig:exp01}
\end{figure}

@@

The residency of the perpetrator is not a significant driver of support for lynching. Subjects were indifferent as to whether the crime perpetrator lives in the area (0.498, SE = 0.004, $p$-value = 0.668) or in another neighborhood (0.502, SE = 0.004, $p$-value = 0.668). In this sense, it is possible that subjects see lynchings not as a means to protect the area in which they live, which people commonly suppose vigilante groups do, but rather as retribution in specific cases.

The next attribute is type of offense. Rape was by far the most significant attribute in our model (0.719, SE = 0.009, $p$-value $<$ 0.001), followed by murder (0.608, SE = 0.006, $p$-value $<$ 0.001) and molestation (0.538, SE = 0.006, $p$-value $<$ 0.001). It is not surprising that murder and rape were the most important offenses to respondents. Overall, we found that respondents do not believe that property crimes, such as car theft (0.351, SE = 0.009, $p$-value $<$ 0.001) and pickpocketing (0.314, SE = 0.006, $p$-value $<$ 0.001), provide justification for lynching. When we analyzed the data by groups, we also saw that poorer respondents, whose household income is less than R\$3,000 (US\$650) per month, are more likely to support the lynching of individuals who commit a property crime.

For age, we observed that respondents were particularly concerned about children as crime victims (0.571, SE = 0.007, $p$-value $<$ 0.001), as this attribute is strongly correlated with support for lynchings. Respondents also thought lynchings are more justified when the victim is a teenager (0.505, SE = 0.007, $p$-value = 0.474), an elderly person (0.479, SE = 0.006, $p$-value $<$ 0.001), or an adult (0.464, SE = 0.006, $p$-value $<$ 0.001).

@@

Lastly, and related to this, we examined who respondents believed are more legitimate perpetrators of lynchings. Respondents considered the family to be the most justified in carrying out a lynching (0.534, SE = 0.007, $p$-value $<$ 0.001). Again, the intervals here did not overlap with those of any other category. Bystanders (0.505, SE = 0.007, $p$-value = 0.450), gangs (0.502, SE = 0.007, $p$-value = 0.815), and neighbors (0.492, SE = 0.007, $p$-value = 0.262) do not increase or decrease respondents attitudes about lynching. If the police carry out a lynching, that reduced the perceived justification (0.467, SE = 0.007, $p$-value $<$ 0.001). In sum, respondents did not believe that lynchings should be carried out by the state but did believe they should be used as a tool for individual or family retribution.

Respondents were given the opportunity to explain the reason why they answered as they did. We analyzed a random sample of nearly 17 percent of the answers. Three major themes emerge. First, respondents felt that crimes that were especially heinous deserved a response. For example, one respondent explained their choice of a particular case as more justified by noting, ``Eye for an eye, it's absolutely fair.'' A second  theme is the injustice of victimizing vulnerable populations. One respondent explains, ``The victim is an elderly person, it's okay to lynch the criminal.'' Another respondent says, ``If the victim is a child, lynchings are not enough, they deserve more.'' Finally, a third theme is that families have a special right to vengeance. One respondent explained their choice by saying, ``The family was appalled and did justice with their own hands!'' These open-ended answers are consistent with the quantitative results about what people believe, but they also provide additional evidence about why they hold their beliefs.
